Coordinated stimulation of axon regenerative and neurodegenerative transcriptional programs by Atf4 following optic nerve injury

Version 1

Abstract

Abstract

Previously we showed that neurodegeneration initiated by axonal insults depends in part on the stress-responsive kinase Perk (Larhammar et al., 2017). Here we show that Perk acts primarily through Activating Transcription Factor-4 (Atf4) to stimulate not only pro-apoptotic but also pro-regenerative responses following optic nerve injury. Using conditional knockout mice, we find an extensive Perk/Atf4-dependent transcriptional response that includes canonical Atf4 target genes and modest contributions by C/ebp homologous protein (Chop). Overlap with c-Jun-dependent transcription suggests interplay with a parallel stress pathway that couples regenerative and apoptotic responses. Accordingly, neuronal knockout of Atf4 recapitulates the neuroprotection afforded by Perk deficiency, and Perk or Atf4 knockout impairs optic axon regeneration enabled by disrupting the tumor suppressor Pten. These findings contrast with the transcriptional and functional consequences reported for CRISPR targeting of Atf4 or Chop and reveal an integral role for Perk/Atf4 in coordinating neurodegenerative and regenerative responses to CNS axon injury.

Introduction

Experiments using conditional knockout mice have revealed a critical role for stress signaling-mediated transcriptional responses in determining the fates of neurons after axon injury. Prominent among these, neuronal knockout of the stress-responsive Dual leucine-zipper kinase (Dlk) suppresses the stimulation by injury of transcription factors of the bZIP family, including c-Jun and the Activating transcription factor-3 (Atf3), that substantially alter the transcriptomes of distressed neurons (Asghari Adib et al. 2018; Farley & Watkins 2018; Le Pichon et al. 2017; Tran et al. 2019; Welsbie et al. 2013, 2017). As is common in cellular stress signaling, this injury-induced MAP kinase (MAPK) signaling cascade couples the activation of transcriptional programs that prime damaged neurons for repair with those that prime those same neurons for apoptosis, the latter a feature that may facilitate the elimination of irreparable cells (Watkins et al. 2013).

These and other essential insights have been enabled by mouse optic nerve crush, an exceptionally tractable model for CNS axon injury that has been valuable for understanding axonopathy-driven CNS neurodegeneration, particularly glaucoma, and the neuron-intrinsic and extrinsic factors that limit CNS axon regeneration (Lindborg et al. 2021; Tran et al. 2019; Vidal-Sanz et al. 2017). Following crush injury, axotomized retinal ganglion cells (RGCs) face both intrinsic and environmental barriers to axon regeneration, and the pro-apoptotic aspects of this response ultimately result in extensive RGC neurodegeneration over subsequent weeks (Syc-Mazurek et al. 2017a; Watkins et al. 2013; Welsbie et al. 2013, 2017). Though neuronal knockout of Dlk, c-Jun, or Atf3 each confer considerable RGC neuroprotection, disrupting these also limits the upregulation of regeneration-associated genes (RAGs) and the success of regenerative interventions, such as knockout of the tumor suppressor Pten, that offer promise for enabling CNS repair (Jacobi et al. 2022; Watkins et al. 2013). Devising strategies to suppress neuronal loss without restricting intrinsic programs for repair will therefore require improving our understanding of which pathways and effectors within the injury response control axon regenerative and neurodegenerative programs and how these overlap (Patel et al. 2020).

Of particular interest is deciphering the contributions of the Integrated Stress Response (ISR). The ISR, activated in parallel to c-Jun as part of the Dlk signaling cascade (Larhammar et al. 2017), results in the suppression of general translation but enhanced transcription and translation of select mRNAs, including Atf4, encoding the Activating transcription factor-4, and Ddit3, encoding the C/ebp homologous protein (Chop) (Pitale et al. 2017). Both Atf4 and Chop have the potential to regulate transcription of Atf3 and/or heterodimerize with c-Jun, Atf3, other transcription factors of the C/ebp family, or each other, raising the possibility of crosstalk between the ISR and MAPK arms of the Dlk response (Pakos-Zebrucka et al. 2016). Genetic disruption of ISR activation or of neuronal Eif2ak3, the gene encoding the ISR-activating kinase Perk, confers substantial RGC neuroprotection after optic nerve crush (Larhammar et al. 2017; Yang et al. 2016). These findings demonstrate that the neuronal ISR represents a functional component of the intrinsic neurodegenerative response but have provided limited mechanistic insight.

Recent studies emphasize the importance of defining the downstream effectors of the ISR, their relationships to c-Jun, and their impacts on RGC axon regenerative potential. Based on CRISPR screening, RNA-seq, and ATAC-seq, Tian et al. (2022) proposed that Atf4 serves, along with C/ebpψ, as a critical component of a core neurodegenerative program, with Chop and Atf3 together mediating a second core program (Tian et al. 2022). Injured RGCs expressing Cas9 and a gRNA pool against Atf4 exhibit improved survival and modulation of genes primarily associated with intrinsic neuronal stressors (e.g., DNA damage response, autophagy, and the NAD/p53 pathway) distinct from canonical Atf4 target genes for adapting to amino acid deprivation, endoplasmic reticulum stress, and other ISR-activating insults. Additive to this, partial neuroprotection could also be achieved by gRNA pools targeting either Ddit3 or Atf3, either of which resulted in extensive, highly overlapping, transcriptional consequences related to cytokines and innate immunity (Tian et al. 2022). Importantly, a companion report shows that, of these same gRNA pools, only targeting Atf3 reduced RGC axon regeneration enabled by knockout of the tumor suppressor Pten (Jacobi et al. 2022). Together, these findings led to the proposal that Atf4 and Chop function in parallel neurodegenerative programs and, unlike Atf3 or c-Jun, do so without influencing RGC axon regenerative potential.

Other reports, however, seem to be inconsistent with these non-canonical roles for Atf4 and Chop. The Dlk-mediated stress response includes ISR-dependent upregulation of Chac1, Ddit3, and other target genes consistent with a typical Atf4-mediated transcriptional program, inclusive of Chop activation (Larhammar et al. 2017). In addition, the proposed role of Chop as a principal neuron-intrinsic effector of the response to optic nerve injury appears to be incongruent with the mild impact of germline Ddit3 knockout on that response. RNA-seq of retinas from these mice uncovered few injury-induced transcripts that were significantly suppressed relative to the extensive impact of neural knockout of c-Jun (Syc-Mazurek et al. 2022). Though these Chop-null mice reproducibly exhibit partial RGC neuroprotection (Hu et al. 2012; Syc-Mazurek et al. 2017b), the modest influence of Chop deficiency on the transcriptome suggests the need for further evaluation of whether its primary role in regulating RGC survival is neuron-autonomous, as it may also reflect developmental effects or influences on other cell types, perhaps outside of the retina.

Here we use previously validated conditional knockout (cKO) mice lines to directly address these and other questions raised by germline knockout and CRISPR studies. We find that a canonical Atf4 response functions as the principal mediator of a Perk-activated response that broadly contributes to both RGC apoptosis and axon regenerative potential. Unexpectedly, neuronal Chop, long considered to be a primary neuron-intrinsic effector of the ISR in injured RGCs, plays a relatively modest role within these Perk/Atf4-stimulated programs. These findings reveal an integral role for Perk-stimulated Atf4 in the response to optic axon damage, serving to link neurodegenerative and axon regenerative transcriptional programs that determine the fates of distressed neurons.

Results

Neuronal Atf4 knockout mimics the neuroprotection provided by Perk deletion

Deletion of neuronal Perk, a key mediator of the ISR after optic nerve crush, provides significant, though incomplete, neuroprotection to RGCs (Larhammar et al. 2017). To determine how the ISR effectors Atf4 and Chop contribute to neurodegeneration, we evaluated the survival of RGCs in cKO mice using immunohistochemistry for the pan-RGC marker RBPMS. Using AAV serotype 2 and the human syapsin-1 promoter (hSyn1) to primarily drive Cre expression in neurons of the ganglion cell layer (Supplemental Figure S1), we observed that targeting Atf4, but, surprisingly, not Ddit3, confers significant neuroprotection of RBPMS-expressing RGCs (Figure 1A). We next compared the neuroprotection measured by this method with that determined using phospho-c-Jun as a robust, easily quantifiable nuclear marker of injured RGCs, finding similar improvements to RGC survival upon Atf4 knockout using either method (Figure 1B). To determine if Chop disruption might augment the neuroprotection conferred by Atf4 disruption, we generated Atf4/Ddit3 double conditional knockout (dcKO) mice. However, these mice did not exhibit significantly greater RGC survival than Atf4 cKO mice (Figure 1C). Finally, we directly compared the effects of Perk knockout and Atf4 knockout, finding disruption of Atf4 is sufficient to afford a similar degree of neuroprotection to that of Perk (Figure 1D). Together with previous findings that Atf4 activation is Dlk- and Perk-dependent (Larhammar et al. 2017), these results suggest that Atf4 serves as the primary mediator of the pro-apoptotic effects of the ISR after optic nerve crush.

Neuronal Atf4 contributes to RGC neurodegeneration after optic nerve crush.

(A) Neuronal knockout of Atf4, but not Ddit3, by intravitreal AAV2-hSyn1-mTagBFP2-ires-Cre in conditional knockout (cKO) mice improves survival of RGCs 14 days after optic nerve crush, as assessed by immunostaining for the pan-RGC marker RBPMS. (B) Neuronal knockout of Atf4 similarly improves survival of RGCs, as assessed by the robust nuclear marker of injured RGCs, phospho-c-Jun (p-c-Jun). (C-D) Comparable levels of neuroprotection are conferred by Atf4 cKO and (C) Atf4/Ddit3 double cKO (dcKO), or (D) Eif2ak3 cKO, which results in deletion of the ISR-activating kinase Perk. Error bars are SEM. Statistical analysis: one-way ANOVA with Tukey’s multiple comparison with a single pooled variance (*p<0.05; ***p<0.001; ****p<0.0001).

Neuronal Perk mediates an extensive contribution to the injury response primarily through Atf4

The equivalent neuroprotection afforded by knockout of Perk and Atf4 raises the possibility that Atf4 may be the principal neuron-autonomous mediator of ISR-activated transcription influencing RGC survival. To better understand the contributions of Perk and its effectors to the transcriptional response, we performed expression profiling of retina by RNA-seq three days after optic nerve crush in wildtype mice and three cKO lines: Eif2ak3 cKO to disrupt Perk, Ddit3 cKO to disrupt Chop, and Atf4 cKO to disrupt Atf4 (Table S1).

We began by determining the contribution of Perk and its effectors to the transcriptional injury response. Using a stringent statistical assessment for differentially expressed genes (DEGs, false discovery rate FDR<0.05), we identified 282 transcripts modulated by injury in control wildtype mice injected with AAV2-hSyn1-Cre. Among these, 117, or 41.5%, reached this same strict threshold for differential expression in a comparison between Perk cKO retinas after injury wildtype retinas after injury (Table S1), with this number increased to 157, or 55.7%, using a moderately relaxed threshold of p<0.01 and FDR<0.2 (Figure 2A). All but three of these change in the opposite direction of their modulation by injury, suggesting a central role for the ISR in mediating these injury responses. Among the 154 transcripts regulated by injury in a Perk-dependent manner, 56 (36.4%) are also suppressed by Atf4 knockout, with a particularly pronounced contribution among those transcripts that are upregulated after injury (37 out of 73, or 50.7%), including seven of the nine transcripts that display Chop-dependence at a threshold of p<0.01 and FDR<0.2 (Figure 2A). Notably, among Atf4-regulated transcripts, 86.2% also exhibit Perk-dependence at this threshold, consistent with Atf4 activation after optic nerve crush being mediated predominantly by Perk and not subject to compensation by other eIF2α kinases (Larhammar et al. 2017). This analysis, though limited by its arbitrary thresholds for defining factor-dependence, implies an essential role of the Perk-activated ISR in the transcriptional injury response, with Chop serving as a relatively minor contributor within a program primarily mediated by Atf4.

The Perk-activated ISR is a prominent contributor to the transcriptional response to optic nerve injury, primarily acting through Atf4.

(A) Venn diagrams of 282 transcripts exhibiting significant differences by RNA-seq between uninjured retinas and 3 days after optic nerve crush (FDR<0.05). Amongst those, transcripts that exhibit differences between wildtype and cKO retinas after injury (p<0.01 and FDR<0.2) were classified as Perk-, Atf4-, or Chop-regulated. (B-D) Linear regression analyses comparing the impact of neuronal Perk deletion (Eif2ak3 cKO), Atf4 deletion (Atf4 cKO), and Chop deletion (Ddit3 cKO) on the same 282 injury-responsive transcripts. Names of representative Atf4-dependent transcripts are burnt orange text and representative Chop-dependent transcripts are gold text. Significant changes are indicated by colored dots (FDR<0.05).

To evaluate these relationships in a threshold-independent manner, we performed linear regression of the 282 injury-regulated genes. We first compared the impact of Perk cKO on the regulation of these transcripts by injury with that of Atf4 cKO. This analysis reveals a strong correlation (R=0.87) that, along with a best-fit slope of 0.75, suggests that the role of Atf4 in mediating the effects of Perk is more comprehensive than suggested by threshold-dependent analyses alone (Figure 2B). A similar assessment uncovers a milder (R=0.59), but still highly significant, correlation between Perk-dependent and Chop-dependent injury-responsive transcripts (Figure 2C). However, the shallow slope of 0.23 for the best-fit line reveals that, though Chop influences similar genes as Perk, it does so relatively mildly. Consistent with these findings, we observe a significant correlation between Atf4-dependent and Chop-dependent transcripts, again with a shallow slope that indicates a more potent influence of Atf4 knockout (Figure 2D). These results imply that Chop serves as a potentiator of the ISR after optic nerve crush rather than a primary mediator. Together, these findings are consistent with Atf4 exerting the dominant role in the transcriptional response and promotion of RGC apoptosis.

Perk signaling regulates canonical Atf4 and c-Jun transcriptional programs

To explore the transcriptional programs downstream of Perk, we next applied the Upstream Regulator tool of Ingenuity Pathway Analysis (IPA), which includes determination of potential transcription factors likely to be mediators of the observed patterns of transcriptional changes and known target genes derived from extensive curation of the literature. Consistent with previous studies (Syc-Mazurek et al. 2022; Yasuda et al. 2016), we find that Atf4, along with c-Jun, emerges among the top five candidates for a positive transcriptional regulator (p<10-13 and Activation Z-score>2.5), based on the upregulation of many genes previously demonstrated to be direct targets of Atf4, including Atf3, Mthfd2, Stc2, and Phgdh (Ben-Sahra et al. 2016; Han et al. 2013; Pan et al. 2007; Zhao et al. 2016) (Figure 3A). IPA also suggested PGC1α (gene name: Ppargc1a), which can be negatively regulated by Atf4 activation (Montori-Grau et al. 2022; Wang et al. 2013), as the most influential negative transcriptional regulator (Table S2). Chop (Ddit3) was implicated among a cluster of dozens of less influential positive regulators (p<10-5 and Activation Z-score>1.25) (Figure 3A; Table S2). Comparisons between wildtype and Perk (Eif2ak3) cKO retinas 3 days after injury suggest that disruption of Perk abrogates Atf4 activation most strongly, with c-Jun and PGC1α also affected (Figure 3B; Table S3). This analysis suggests the activation of a canonical Perk/Atf4-mediated transcriptional response (Figure 3C) after RGC axon injury and raised the unexpected possibility that Perk signaling might also influence c-Jun-regulated target genes.

Perk regulates RGC-autonomous transcriptional changes through canonical Atf4 target genes and influence on c-Jun-regulated programs.

(A-B) Volcano plots of Ingenuity Pathway Analysis (IPA) implicate Atf4 and c-Jun as potential Upstream Transcriptional Regulators of injury-induced expression changes in (A) wildtype retina, and (B) those whose activity after injury is reduced by neuronal Perk deletion. Red orange, blue, and light blue dots indicate Z>2.5, p<10-13; Z>1.5, p<10-10; Z<-2.5, p<10-13; and Z<-1.5, p<10-10, respectively. (C) Heat map of RNA-seq data, showing known and putative Atf4 target genes after injury in wildtype or cKO retinas. (D-F) Linear regression analyses comparing current whole retina RNA-seq cKO data to published data from a similarly designed experiment (Syc-Mazurek et al, 2022). Data from that independent study is indicated on the y-axis by &. (D) Strong correlation among 282 injury-responsive transcripts in the respective wildtype injury conditions between these two independent studies. Names of representative Atf4-dependent transcripts are burnt orange text and representative Chop-dependent transcripts are gold text. (E-F) Significant correlations between the impacts of (H) neuronal Perk, or (I) Atf4 conditional knockout (this study) and deletion of c-Jun from the majority of neural retina& (p<0.001). (J) Cross-study comparison of Ddit3 cKO (this study) and Ddit3 KO& shows strong correlation between the few Chop-dependent injury-responsive transcripts (red dots). (G) Volcano plot of retinal expression data three days after optic nerve crush (this study) for 597 transcripts that are significantly altered in RGCs at day 2 or day 4 after crush, as determined by scRNA-seq (Tran et al., 2019). 287 of these transcripts are detected at FDR<0.05 (red, upregulated, or blue, downregulated) or |log2FC|>0.25 (light red, upregulated, or light blue, downregulated) in whole retina. Only three transcripts (yellow dots) were significantly regulated in whole retina in the discordant direction from scRNA-seq findings. (H) scRNA-seq data set (Tran et al., 2019) reveals RGC-autonomous activation of canonical ISR transcripts. Dot plot at 2 and 4 days after injury, showing both pan-RGC and type-specific upregulation of transcripts demonstrated by whole retina RNA-seq to be Atf4- and/or Chop-dependent. Notably, some transcripts exhibiting low, non-uniform, injury induction in RGCs – Avil, Gm29374, Cdsn, and Fibin – were amongst those that appear to exhibit potential Chop-dependence but failed to reach threshold for inclusion in either retinal injury-dependence, Chop-dependence, or scRNA-seq pseudo-bulk analyses. Conversely, Stk32a, the lone Chop-dependent transcript reported by Ddit3 knockout but not neuronal Ddit3 cKO is not regulated by injury within RGCs, suggesting non-autonomous modulation of this transcript by Chop in retina.

To further investigate a potential interaction between c-Jun- and Perk-regulated transcripts, we leveraged RNA-seq datasets from a separate, similarly designed, study that included mice lacking c-Jun in the majority of neural retina and Chop-null mice (Syc-Mazurek et al. 2022). First, we evaluated the expression profiles of the wildtype injury conditions between these independent studies, finding an exceptional correlation (R=0.95, Slope = 1.07) among 282 injury-regulated genes (Figure 3D). That robust relationship between the control conditions provided a basis for further comparisons. We therefore proceeded with a cross-study analysis comparing the impact of Perk cKO with that of c-Jun cKO. That analysis revealed a significant correlation (R=0.49; Slope = 0.44) among c-Jun- and Perk-modulated transcripts (Figure 3E). Nevertheless, some injury-modulated transcripts are regulated exclusively by c-Jun (e.g., Sox11 and Arid5a) or Perk (e.g., Chac1 and Phgdh). Together, these findings suggest that, despite a lack of influence of Jun knockout on Eif2ak3, Atf4, or Ddit3 transcripts and vice versa, these parallel pathways participate in substantially overlapping transcriptional programs.

We next evaluated the similarity between germline and neuronal knockout of Ddit3 between these two studies. Though each uncovered only a small number of Chop-dependent, injury-responsive transcripts, those few exhibit considerable overlap. Just four transcripts passed a stringent test for Chop-dependence in retinas of Ddit3-/- mice (Syc-Mazurek et al. 2022), three of which – Stbd1, Gm13889, and Avil (also known as Doc6 for “dependent on Chop-6”) – have been demonstrated to be upregulated neuron-autonomously by injury within RGCs by scRNA-seq (Tran et al. 2019). Using similar stringency (FDR<0.05), these three were among only five genes that we independently detected as injury-upregulated in a neuronally Chop-dependent manner using Ddit3 cKO mice (Figure 3F). Additional injury-induced transcripts suppressed by greater than 50% in both Ddit3 cKO and Ddit3 KO (though without surpassing the FDR<0.05 threshold) include Fibin (p<0.0005 in both studies) and Cdsn (p<0.001 in both studies), genes for which ChIP-seq has previously identified Chop binding near their transcriptional start sites under stress conditions (Han et al. 2013). These results show that, despite differences in neuroprotection between Ddit3-/- mice (Hu et al. 2012; Syc-Mazurek et al. 2017b) and neuronal targeting of Ddit3 in cKO retinas, their transcriptional consequences within the retina closely resemble one another and align with expected Chop target genes.

Together, these expression profiles suggest that the Perk-mediated ISR is an integral contributor to the retinal transcriptional response after optic nerve injury, acting primarily through canonical Atf4 targets and interacting with c-Jun-mediated transcriptional programs.

RGC-autonomous Atf4- and Chop-dependent transcriptional changes are prominently represented by whole retina transcriptomics

Our targeted expression of Cre recombinase in cKO mice argues for a central role for the neuronal ISR in the response to optic axon injury, with whole retina transcriptomics likely reporting both these neuron-autonomous effects and secondary consequences to other retinal cells. To determine the extent to which ISR-dependent transcription uncovered by retinal RNA-seq represents RGC-autonomous expression changes, we next leveraged an independent single-cell RNA-seq (scRNA-seq) data set that details the injury-induces expression changes within subtypes of RGCs at 2 and 4 days after optic nerve crush (Tran et al. 2019).

We began by assessing the whole retina expression of transcripts that have been demonstrated to be differentially expressed by RGCs at 2 or 4 days after optic nerve crush (pseudo-bulk scRNA-seq |Diff>0.3| and FDR<10-20). Of 615 such transcripts, 597 are represented in wildtype whole retina 3 days after injury. We find that 290 of these transcripts exhibit differential expression (|log2FC|>0.25) in retina, with all but three modulated by injury in the same direction as reported by scRNA-seq (Figure 3G). Moreover, the majority of the 282 injury-responsive transcripts we have identified in whole retina exhibit concordance with changes found in scRNA-seq of RGCs (Supplemental Figure S2A-C). Importantly, many transcripts identified as Perk-, Atf4-, and/or Chop-dependent in this and other studies are among those that are upregulated neuron-autonomously within multiple subtypes of RGCs following optic nerve crush (Figure 3G). Consistent with this, the relationships we detected between Perk-dependent transcripts and Atf4- or c-Jun-dependent transcripts in whole retina are maintained when re-assessed using DEGs confirmed to be injury-regulated within RGCs by scRNA-seq (Supplemental Figure S2D-E). These results imply that the injury-induced Atf4 transcriptional program that we detect by whole retina transcriptomics of cKO mice primarily reflects RGC-autonomous expression changes.

This canonical Atf4 program, inclusive of the contribution of Chop, contrasts with a recent report of two non-canonical, largely non-overlapping, programs controlled by these two ISR transcription factors (Tian et al. 2022). We next investigated potential sources of this discordance. The proposed parallel neurodegenerative programs were deduced in part by RNA-seq of FACS-enriched injured RGCs expressing gRNAs targeting Atf4 or Ddit3. We therefore began with an assessment of how the injury responses detected in whole retina in the current study compare with those reported for RGCs collected by FACS. Using only transcripts independently identified as RGC-autonomous expression changes by scRNA-seq (Tran et al. 2019) for this cross-study comparison, we find a highly significant correlation (R=0.74) between injury-regulated transcripts under control conditions (i.e., without gene targeting) between these independent studies (Supplemental Figure S3A). This suggests that these distinct approaches similarly report prominent RGC-autonomous expression changes, inclusive of numerous known direct targets of Atf4, with a steep linear regression (slope = 1.57) that, unsurprisingly, suggests greater sensitivity for these changes using FACS-sorted RGCs. Despite that similarity, linear regression reveals little correlation between the impact of cKO- and gRNA-mediated knockout on 597 RGC-autonomous, injured-regulated transcripts. We find only very weak, shallow correlation between the effects of Atf4 gRNA and Atf4 cKO (R=0.13, slope = 0.33, p<0.005). cKO-sensitive signature Atf4 target genes Chac1, Phgdh, Sesn2, Slc7a5, and Slc6a8, amongst others, are not among the 125 of these 597 (20.9%) transcripts reported to be significantly affected by Atf4 gRNA (Supplemental Figure S3B) (Crawford et al. 2015; Garaeva et al. 2016; Han et al. 2013; Liu et al. 2018; Oh-Hashi et al. 2013; Park et al. 2017). We also find no correlation between the influence of Ddit3 cKO and gRNA targeting Ddit3, with no reported inhibition by this gRNA pool of the Ddit3 KO- and cKO-sensitive transcripts Gm13889 and Stbd1 among 121 (20.3%) significantly affected transcripts (FDR<0.05) (Supplemental Figure S3C). Notably, the reported effects of gRNA targeting Ddit3, like Atf3 gRNA, include no reduction of Ddit3 transcript but greater than 80% suppression of the pro-apoptotic and pro-regenerative Atf3 transcript (Tian et al. 2022), an effect not observed in Ddit3 germline (Syc-Mazurek et al. 2022) or conditional knockout mice (present study), despite ample sensitivity for Atf3 modulation using whole retina transcriptomics. These analyses argue that more specific and effective disruption of target genes by cKO primarily accounts for the striking discordance between the transcriptional programs of Atf4 and Chop suggested by these distinct approaches.

Neuronal Atf4 knockout limits axon regeneration by Pten-deficient RGCs

The commonalities that we uncovered between c-Jun- and Perk-regulated transcription suggested the hypothesis that, like c-Jun, Perk/Atf4 could be involved not only in promoting apoptosis but also in promoting regeneration. To investigate this possibility, we first examined the Perk- and Atf4-dependence of regeneration-associated genes (RAGs) and the injury-induced downregulation of mature and subtype-specific RGC markers, finding that many exhibit at least partial Perk- and Atf4-dependence (Figure 4A-B). We therefore crossed mice harboring the floxxed Pten alleles and those harboring floxxed Atf4 alleles or floxxed Perk alleles to generate homozygous dcKO mice. Following intravitreal injection of AAV2-hSyn1-Cre, we performed optic nerve crush, comparing axon regeneration in these dcKO mice to that of Pten cKO mice (Figure 4C-D). We find that neuronal Perk deficiency or Atf4 deficiency limits the efficacy of this regenerative intervention. These data support an integral contribution of the ISR to the axon regenerative program in these CNS neurons, and indicate that, as with Dlk and c-Jun, Atf4 is involved in both pro-regenerative and pro-apoptotic transcriptional changes.

Perk-Atf4 contributes to RGC axon regenerative potential after optic nerve crush.

(A-B) Heat maps of select injury-responsive transcripts revealed by retinal RNA-seq three days after optic nerve crush in wildtype and cKO mice, focusing on (A) genes implicated in axon regeneration, and (B) the maintenance of mature and subtype-specific RGC phenotypes. (C-D) Double conditional knockout (dcKO) of neuronal Perk (Eif2ak3; n=4) or Atf4 (Atf4; n=7) reduces regeneration enabled compared to conditional knockout (cKO) of the tumor suppressor Pten alone (n=8). Error bars are SEM. Statistical analysis: two-way ANOVA with Sidak’s multiple comparisons test (**p<0.01).

Discussion

Cellular stress signaling pathways typically couple the promotion of growth and repair with priming for apoptosis, a feature of stress responses that may aid in the elimination of cells that are irretrievably damaged (Hotamisligil & Davis 2016). As prominent effectors of stress signaling, transcription factors of the bZIP family, including c-Jun and Atf3, therefore can contribute either to apoptosis or to recovery, depending on context, complicating efforts to harness transcriptional programs for axon regeneration independently of neurodegenerative programs (Jacobi et al. 2022; Kole et al. 2020; Raivich et al. 2004; Simon & Watkins 2018; Syc-Mazurek et al. 2017b; Tian et al. 2022; Watkins et al. 2013). Multiple lines of evidence indicating that the Perk-activated ISR contributes to RGC apoptosis after optic nerve crush raised the appealing possibility that this pathway may represent a distinct branch of the injury response that might be targeted to selectively reduce neurodegeneration without suppressing repair programs (Hu et al. 2012; Larhammar et al. 2017; Tian et al. 2022; Wang et al. 2020). The cKO studies reported here, however, reveal that the Perk-activated ISR is more integral to the entire RGC injury response than previously appreciated, contributing substantially, in conjunction with c-Jun, to both apoptotic and regenerative programs. Though MAP kinase stress signaling and the ISR are known to exhibit some degree of crosstalk in other settings (Brown et al. 2016; Danzi et al. 2018; Pakos-Zebrucka et al. 2016), the exceptionally close coupling of these largely independent pathways by Dlk signaling seems to be a critical feature of the response of RGCs to axon injury, with disruption of Atf4, like that of c-Jun, limiting both neurodegeneration and axon regenerative potential.

By examining in parallel the neuronal knockout of Perk and its effectors Atf4 and Chop, along with comparisons to related studies, the current work has provided insights into the mechanisms by which the ISR influences the fates of injured RGCs. These conditional knockout mouse experiments reveal that instead of non-canonical Atf4 and Chop transcriptional programs, each with distinct contributions to neurodegeneration, Atf4 is the principal effector of the ISR transcriptional response, with Chop playing a modest role within a canonical Atf4-mediated program. Though RGC neuroprotection in germline Ddit3 knockout mice has long been interpreted as evidence of its central role in the neuronal ISR, our findings suggest that further investigations of Chop may instead uncover its non-autonomous mechanisms regulating the survival of injured RGCs. The unexpectedly subtle role for neuronal Chop aligns well with expression profiling data from Chop-null mice (Syc-Mazurek et al. 2022) but stands in contrast to the extensive impact of CRISPR-based targeting of Ddit3 on the RGC transcriptome, which includes potent suppression of Atf3 and mimics with remarkable precision (R>0.9, slope=1) the impact of Atf3 gRNA (Tian et al. 2022). It is possible then that the differential impact between Ddit3 cKO and Ddit3-targeting gRNA on Atf3 mRNA underlies their distinct effects on transcription and neuroprotection, though this hypothesis is not sufficient to clarify why this Atf3-suppressing Ddit3 gRNA pool does not recapitulate the inhibition of RGC axon regeneration by Atf3 gRNA (Jacobi et al. 2022). With cKO experiments revealing that Atf3 upregulation is dependent on Atf4 and c-Jun rather than Chop, it will be of interest to elucidate how much of the commonality among the transcriptional programs of these bZIP factors is attributable to interdependence of their expression and how much to their combinatorial heterodimerization. Among other non-exclusive possibilities, the commonalities between Atf4- and c-Jun-dependent transcription could reflect: (1) dependence of Atf3 elevation on c-Jun-Atf4 heterodimers; (2) dependence on Atf4 for induction of Atf3, which then heterodimerizes with c-Jun; or (3) regulation of common target genes through distinct DNA binding sites. The current study provides a framework for deciphering these networks in part by leveraging cross-study comparisons of retina, RGC, and single-cell RNA-seq data to uncover robust RGC-autonomous expression signatures of Atf4 and other transcription factors.

Given the potential for the ISR to influence neuronal fate through multiple translational and transcriptional mechanisms, it is perhaps surprising that its primary contributions to the fates of injured RGCs are mediated by a single transcription factor, Atf4. Nevertheless, at least four lines of evidence contend that the cKO experiments reported here provide a reliable picture of its contribution. First, the extensive commonality in injury-regulated transcripts across studies, whether utilizing retinal expression data or filtering for RGC-autonomous expression changes, argues against AAV2-hSyn1-Cre or other technical aspects of our approach altering the injury response. Secondly, support for the modest impact of Chop disruption on RGC injury-regulated genes is provided by the highly similar findings in the independently generated and validated Ddit3-null and Ddit3 cKO mouse lines. Third, we find remarkable overlap between expression changes and phenotypes mediated by Perk and Atf4, which we previously demonstrated to be in the same pathway after optic nerve crush and other Dlk-activating insults (Larhammar et al. 2017). Finally, the interaction between Perk/Atf4 and c-Jun-regulated transcription uncovered by our cross-study analysis predicts an impact of Perk or Atf4 knockout on RGC axon regenerative potential that is validated by dcKO with Pten. This result is consistent with recent findings that early injury responses, including regenerative and degenerative responses that we have found in this study to be driven by Perk/Atf4, are similar in non-regenerating wildtype and regenerating Pten-deficient RGCs (Jacobi et al, 2022). Together, these conditional knockout studies highlight the critical roles for ISR-activated transcriptional programs in determining the fates of distressed neurons and the intimate link between injury-activated neurodegenerative and axon regenerative programs.

Supporting information

eLife Assessment

This study is important as it examined the role of Perk (Protein kinase RNA-like endoplasmic reticulum kinase) and Atf4 (Activating Transcription Factor-4) in the integrated neurodegenerative and regenerative responses following the optic nerve injury. The evidence is convincing as the authors relied on both newly generated transcriptomic data and publicly available databases to support their research. While there are some limitations in data quality and interpretations, the study is likely to be of interest to researchers studying optic neuropathies and axonal regeneration.

Reviewer #1 (Public Review):

Somasundaram and colleagues explore the role of transcription factors in retinal ganglion cell (RGC) death and axonal regeneration after a disease relevant insult (mechanical axonal injury). The work significantly extends our knowledge of the role of MAPK and integrated stress response (ISR) in controlling RGC fate after injury. Specifically, the manuscript shows that after axonal injury PERK-activated ISR acts through Atf4 to drive a prodeath transcriptional response in RGCs, in part by crosstalk with the prodeath JUN transcriptional program. Also, and perhaps most interesting, the work shows that PERK-ATF4 pathway activation is pro-regenerative for RGC axons. A major plus of the manuscript is that many new RNA-seq datasets are generated that describe the major prodegenerative and proregenerative gene networks altered after axonal injury.

A limitation of the study is that it does not directly compare the effect of inhibiting the PERK-ATF4 pathway with inhibiting JUN and/or JUN-CHOP double deficient animals. It would also be useful, for the cell survival experiments shown in Figure 1, to examine a longer time point than 14 days to understand the long-term consequence of manipulating the PERK-ATF4 pathway.

Reviewer #2 (Public Review):

This manuscript investigates the role of Perk (Protein kinase RNA-like endoplasmic reticulum kinase) and Atf4 (Activating Transcription Factor-4) in neurodegenerative and regenerative responses following optic nerve injury. The authors employed conditional knockout mice to examine the impact of the Perk/Atf4 pathway on transcriptional responses, with a particular focus on canonical Atf4 target genes and the involvement of C/ebp homologous protein (Chop).

The study demonstrates that Perk primarily operates through Atf4 to stimulate both pro-apoptotic and pro-regenerative responses after optic nerve injury. This Perk/Atf4-dependent response encompasses canonical Atf4 target genes and limited contributions from Chop, exhibiting overlap with c-Jun-dependent transcription. Consequently, the Perk/Atf4 pathway appears crucial for coordinating neurodegenerative and regenerative responses to central nervous system (CNS) axon injury. Additionally, the authors observed that neuronal knockout of Atf4 mimics the neuroprotection resulting from Perk deficiency. Moreover, Perk or Atf4 knockout hinders optic axon regeneration facilitated by the deletion of the tumor suppressor Pten.

These findings contrast with the transcriptional and functional outcomes reported for CRISPR targeting of Atf4 or Chop, revealing a vital role for the Perk/Atf4 pathway in orchestrating neurodegenerative and regenerative responses to CNS axon injury.

However, the main concern is the overall data quality, which appears to be suboptimal. The transfection efficiency of AAV2-hSyn1-mTagBFP2-ires-Cre used in this study does not seem highly effective, as evidenced by the data presented in Supplementary Figure 1. The manuscript also contains several inconsistencies and a mix of methods in data collection, analysis, and interpretation, such as the labeling and quantification of RGCs and the combination of bulk and single-cell sequencing results.

Despite these limitations, the study offers valuable insights into the role of the Perk/Atf4 pathway in determining neuronal fate after axon injury, emphasizing the significance of understanding the molecular mechanisms that govern neuronal survival and regeneration. This knowledge could potentially inform the development of targeted therapies to promote neuroprotection and CNS repair following injury.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public Review):

A limitation of the study is that it does not directly compare the e4ect of inhibiting the PERKATF4 pathway with inhibiting JUN and/or JUN-CHOP double deficient animals. It would also be useful, for the cell survival experiments shown in Figure 1, to examine a longer time point than 14 days to understand the long-term consequence of manipulating the PERK-ATF4 pathway.

We appreciate that both suggestions are fantastic ideas for future studies but consider them to be beyond the scope of this investigation.

Reviewer #2 (Public Review):

However, the main concern is the overall data quality, which appears to be suboptimal. The transfection e4iciency of AAV2-hSyn1-mTagBFP2-ires-Cre used in this study does not seem highly e4ective, as evidenced by the data presented in Supplementary Figure 1.

We appreciate the importance of the e;ectiveness of transfection e;iciency of AAV2-hSyn1-mTagBFP2-ires-Cre to the interpretations of our results and acknowledge that the imaging and color schemes used required improvement. We have now validated widespread knockout in RGCs using AAV2-hSyn1-mTagBFP2-ires-Cre, improving the staining and imaging of LSL.tdTomato Cre reporter mice (Figure S1A-B) and using RNAScope to validate the disruption of ATF4 and CHOP, respectively, in the RGCs of ATF4 cKO and CHOP cKO mice (Figure S1C-D). Additional validation of functional knockout of these transcription factors is provided by reduction of RGC-autonomous expression of transcripts that we identified in this study to be injury-regulated in an ATF4-dependent (Chac1, Atf3, Figure 4C-E) or ATF4- and CHOP-dependent manner (Ecel1, Avil, Figure 4C-E and Figure S2D).

The manuscript also contains several inconsistencies and a mix of methods in data collection, analysis, and interpretation, such as the labeling and quantification of RGCs and the combination of bulk and single-cell sequencing results.

Regarding the use and comparison of bulk-seq and scRNA-seq data, it is our sense that these innovative approaches will be among the impactful aspects of this study. Numerous transcriptomic studies of the optic nerve crush model exist, though it has been unclear whether major and minor technical di;erences would preclude deriving insights across studies without the expense and time of exact reproduction. One goal of this study was to evaluate the hypothesis that, despite the obvious limitation that RGCs represent fewer than 1% of cells in whole retina bulk transcriptomics approach, the signals amongst top di;erentially expressed genes (DEGs) would be dominated by injury-induced changes within RGCs and that the most robust of these changes would be readily detected across techniques and labs, serving as a cornerstone for interpreting similarities and di;erences in findings. We believe that the results validate this approach. Important insights gained in this study from these cross-study and cross-platform analyses include:

(1) Genes that we identify in this study as neuronal ATF4-dependent by whole retina transcriptomics include many of the most robust genes expression changes observed across multiple studies that enrich for RGCs and those that only report RGC-autonomous expression changes by scRNA-seq. This observation predicts that many of the ATF4-dependent expression changes that we report are RGC autonomous, which we further validate in this revision by RNAScope.

(2) Similarly designed whole transcriptomics studies across labs can be remarkably robust for top DEGs, showing striking similarity that allows for meaningful insights and testable hypotheses across di;erent knockout and conditional knockout mice.

(3) scRNA-seq of RGCs and bulk sequencing of FACS-enriched RGCs, unsurprisingly results in higher sensitivity for injury-induced expression changes, but the high degree of similarity that we demonstrate between the top DEGs from those studies and whole retina transcriptomics studies allows for confident inferences regarding the expected cell autonomy of reported expression changes in this model, using available resources such as the Single Cell Portal, without the expense and technical optimization required for extensive spatial transcriptomics across numerous mouse models.

Other revisions

In addition to these updates to address the public reviews, we are grateful for the reviewers’ additional recommendations and provide these further revisions:

(1) We appreciate the request to clarify with a schematic the di;erences between our study and a previous report (Tian et al., 2022). A second Correction to that study was published in July 2024, resulting in changes to the logFC values used in our original cross-study comparison and adjustments to multiple figures and tables related to the proposed transcriptional programs of ATF4, CHOP, and the other purported core transcription factors. We have therefore updated our Figure S3A-C in accordance with that Correction to better reflect the underlying data of that study. These changes do not alter our original conclusions that: (a) both the whole retina transcriptomics approach of our study and the FACS-enriched RGC approach of that study readily detect the strong upregulation of many known ATF4 target genes after optic nerve crush (Figure S3A); and (b) there are striking di;erences in the ATF4- and CHOP-dependent transcripts suggested by our cKO data and those suggested by the reported gRNA data. Though we had hoped that the Correction would allow us in this revision to diagram those findings and model for comparison to these cKO findings, documenting those changes and their impacts on the proposed model is beyond the scope of this study.

(2) We agree that the discordance between the gene and protein names for Ddit3/CHOP and Eif2ak3/PERK represents a challenge for clarity, even when gene names are carefully selected when referring to genes or transcripts and protein names when referring to proteins. We have therefore attempted to streamline the naming throughout, using where possible both names.

Version 2

Abstract

Abstract

Stress signaling is important for determining the fates of neurons following axonal insults. Previously we showed that the stress-responsive kinase PERK contributes to injury-induced neurodegeneration (Larhammar et al., 2017). Here we show that PERK acts primarily through Activating Transcription Factor-4 (ATF4) to stimulate not only pro-apoptotic but also pro-regenerative responses following optic nerve damage. Using conditional knockout mice, we find an extensive PERK/ATF4-dependent transcriptional response that includes canonical ATF4 target genes and modest contributions by C/EBP Homologous Protein (CHOP). Overlap with c-Jun-dependent transcription suggests interplay with a parallel stress pathway that orchestrates regenerative and apoptotic responses. Accordingly, neuronal knockout of ATF4 recapitulates the neuroprotection afforded by PERK deficiency, and PERK or ATF4 knockout impairs optic axon regeneration enabled by disrupting the tumor suppressor PTEN. These findings reveal an integral role for PERK/ATF4 in coordinating neurodegenerative and regenerative responses to CNS axon injury.

Introduction

Experiments using conditional knockout mice have revealed a critical role for stress signaling-mediated transcriptional responses in determining the fates of neurons after axon injury. Prominent among these, neuronal knockout of the stress-responsive Dual leucine-zipper kinase (DLK) suppresses the stimulation by injury of transcription factors of the bZIP family, including c-Jun and the Activating transcription factor-3 (ATF3), that substantially alter the transcriptomes of distressed neurons (Asghari Adib et al., 2018; Farley and Watkins, 2018; Le Pichon et al., 2017; Tran et al., 2019; Welsbie et al., 2017, 2013). As is common in cellular stress signaling, this injury-induced MAP kinase (MAPK) signaling cascade couples the activation of transcriptional programs that prime damaged neurons for repair with those that prime those same neurons for apoptosis, the latter a feature that may facilitate the elimination of irreparable cells (Watkins et al., 2013).

These and other essential insights have been enabled by mouse optic nerve crush, an exceptionally tractable model for CNS axon injury that has been valuable for understanding axonopathy-driven CNS neurodegeneration, particularly glaucoma, and the neuron-intrinsic and extrinsic factors that limit CNS axon regeneration (Lindborg et al., 2021; Tran et al., 2019; Vidal-Sanz et al., 2017). Following crush injury, axotomized retinal ganglion cells (RGCs) face both intrinsic and environmental barriers to axon regeneration, and the pro-apoptotic aspects of this response ultimately result in extensive RGC neurodegeneration over subsequent weeks (Syc-Mazurek et al., 2017a; Watkins et al., 2013; Welsbie et al., 2013, 2017). Though neuronal knockout of DLK, c-Jun, or ATF3 each confer considerable RGC neuroprotection, disrupting these also limits the upregulation of regeneration-associated genes (RAGs) and the success of regenerative interventions, such as knockout of the tumor suppressor PTEN, that offer promise for enabling CNS repair (Jacobi et al., 2022; Watkins et al., 2013). Devising strategies to suppress neuronal loss without restricting intrinsic programs for repair will therefore require improving our understanding of which pathways and effectors within the injury response control axon regenerative and neurodegenerative programs and how these overlap (Patel et al., 2020).

Of particular interest is deciphering the contributions of the Integrated Stress Response (ISR). The ISR, activated in parallel to c-Jun as part of the DLK signaling cascade (Larhammar et al., 2017), results in the suppression of general translation but enhanced transcription and translation of select mRNAs, including Atf4, encoding the Activating transcription factor-4 (ATF4), and Ddit3, encoding the C/EBP homologous protein (CHOP) (Pitale et al., 2017). Both ATF4 and CHOP have the potential to regulate transcription of ATF3 and/or heterodimerize with c-Jun, ATF3, other transcription factors of the C/EBP family, or each other, raising the possibility of crosstalk between the ISR and MAPK arms of the DLK response (Pakos-Zebrucka et al., 2016). Genetic disruption of ISR activation or of neuronal Eif2ak3, the gene encoding the ISR-activating kinase PERK, confers significant but incomplete RGC neuroprotection after optic nerve crush (Larhammar et al., 2017; Yang et al., 2016). These findings demonstrate that the neuronal ISR represents a functional component of the intrinsic neurodegenerative response but have provided limited mechanistic insight.

Recent studies emphasize the importance of defining the downstream effectors of the ISR, their relationships to c-Jun, and their impacts on RGC axon regenerative potential. Based on CRISPR screening, RNA-seq, and ATAC-seq, Tian et al. (2022) proposed that ATF4 serves, along with C/EBPψ, as a critical component of a core neurodegenerative program, with CHOP and ATF3 together mediating a second core program (Tian et al., 2022). Injured RGCs expressing Cas9 and a gRNA pool targeting ATF4 exhibited improved survival and modulation of genes primarily associated with intrinsic neuronal stressors (e.g., DNA damage response, autophagy, and the NAD/p53 pathway) distinct from canonical ATF4 target genes for adapting to amino acid deprivation, endoplasmic reticulum stress, and other ISR-activating insults. Additive to this, partial neuroprotection could also be achieved by gRNA pools targeting either Ddit3/CHOP or ATF3, either of which resulted in extensive, highly overlapping, transcriptional consequences related to cytokines and innate immunity (Tian et al., 2024, 2022). Importantly, a companion report shows that, of these same gRNA pools, only targeting ATF3 reduced RGC axon regeneration enabled by knockout of the tumor suppressor PTEN (Jacobi et al., 2022). Conditional knockout of ATF4 and CHOP has been reported to provide more neuroprotection than either alone, despite knockout of either appearing in that study to be sufficient to deplete the mRNA for both (Fang et al., 2023). Together, these findings led to the proposal that ATF4 and CHOP function in parallel neurodegenerative programs and, unlike ATF3 and c-Jun, do so without influencing RGC axon regenerative potential.

Other reports, however, seem to be inconsistent with these non-canonical roles for ATF4 and CHOP. The DLK-mediated stress response includes ISR-dependent upregulation of Chac1, Ddit3, and other target genes consistent with a typical ATF4-mediated transcriptional program, inclusive of CHOP activation (Larhammar et al., 2017). In addition, the proposed role of CHOP as a principal neuron-intrinsic effector of the response to optic nerve injury appears to be incongruent with the mild impact of germline Ddit3/CHOP knockout on that response. RNA-seq of retinae from these mice uncovered few injury-induced transcripts that were significantly suppressed relative to the extensive impact of neural knockout of c-Jun (Syc-Mazurek et al., 2022). Though these CHOP-null mice reproducibly exhibit partial RGC neuroprotection (Hu et al., 2012; Syc-Mazurek et al., 2017b), the modest influence of CHOP deficiency on the transcriptome suggests the need for further evaluation of whether its primary role in regulating RGC survival is neuron-autonomous, as it may also reflect developmental effects or influences on other cell types, perhaps outside of the retina.

Here we use previously validated conditional knockout (cKO) mice lines to directly address these and other questions raised by germline knockout and CRISPR studies. We find that a canonical ATF4 response functions as the principal mediator of a PERK-activated response that broadly contributes to both RGC apoptosis and axon regenerative potential. Unexpectedly, neuronal CHOP, long considered to be a primary neuron-intrinsic effector of the ISR in injured RGCs, plays a relatively modest role within these PERK/ATF4-stimulated programs. These findings reveal an integral role for PERK-stimulated ATF4 in the response to optic axon damage, serving to link neurodegenerative and axon regenerative transcriptional programs that determine the fates of distressed neurons.

Results

Neuronal ATF4 knockout mimics the neuroprotection provided by PERK deletion

Deletion of neuronal PERK, a key mediator of the ISR after optic nerve crush, provides significant, though incomplete, neuroprotection to RGCs (Larhammar et al., 2017). To determine how the ISR effectors ATF4 and CHOP contribute to neurodegeneration, we evaluated the survival of RGCs in cKO mice using immunohistochemistry for the pan-RGC marker RBPMS. We used high-titer intravitreal injection of AAV serotype 2 (AAV2) harboring the human synapsin-1 promoter (hSyn1), resulting in widespread Cre activity throughout neurons of the ganglion cell layer and inner nuclear layer of the retina (Supplemental Figure S1A-B). We confirmed the expected loss of Atf4 transcript in ATF4 cKO mice and Ddit3/CHOP transcript in CHOP cKO mice by RNAScope (Supplemental Figure S1C-D). Evaluation of neuronal survival two weeks after optic nerve crush revealed that targeting ATF4, but, surprisingly, not CHOP, confers significant neuroprotection of RBPMS-expressing RGCs (Figure 1A). We next compared the neuroprotection measured by this method with that determined using phospho-c-Jun as a robust, easily quantifiable nuclear marker of injured RGCs, finding similar improvements to RGC survival upon ATF4 knockout using either method (Figure 1B). To determine if CHOP disruption might augment the neuroprotection conferred by ATF4 disruption, we generated ATF4/CHOP double conditional knockout (dcKO) mice. However, these mice did not exhibit significantly greater RGC survival than ATF4 cKO mice (Figure 1C). Finally, we directly compared the effects of PERK knockout and ATF4 knockout, finding disruption of ATF4 is sufficient to afford a similar degree of neuroprotection to that of PERK (Figure 1D). Together with previous findings that ATF4 activation is DLK- and PERK-dependent (Larhammar et al., 2017), these results suggest that ATF4 serves as the primary mediator of the pro-apoptotic effects of the ISR after optic nerve crush.

Neuronal ATF4 contributes to RGC neurodegeneration after optic nerve crush.

(A) Neuronal knockout of ATF4, but not CHOP (Ddit3), by intravitreal AAV2-hSyn1-mTagBFP2-ires-Cre in conditional knockout (cKO) mice improves survival of RGCs 14 days after optic nerve crush, as assessed by immunostaining for the pan-RGC marker RBPMS. (B) Neuronal knockout of ATF4 similarly improves survival of RGCs when assessed by the robust nuclear marker of injured RGCs, phospho-c-Jun (p-c-Jun). (C-D) Comparable levels of neuroprotection are conferred by ATF4 cKO and (C) ATF4/CHOP double cKO (dcKO), or (D) Eif2ak3 cKO, which results in deletion of the ISR-activating kinase PERK. Error bars are SEM. Statistical analysis: one-way ANOVA with Tukey’s multiple comparison with a single pooled variance (*p<0.05; ***p<0.001; ****p<0.0001).

Neuronal PERK mediates an extensive contribution to the injury response primarily through ATF4

The equivalent neuroprotection afforded by knockout of PERK and ATF4 raises the possibility that ATF4 may be the principal neuron-autonomous mediator of ISR-activated transcription influencing RGC survival. To better understand the contributions of PERK and its effectors to the transcriptional response, we performed expression profiling of retina by RNA-seq three days after optic nerve crush in wildtype mice and three cKO lines: Eif2ak3 cKO to disrupt PERK, Ddit3 cKO to disrupt CHOP, and Atf4 cKO to disrupt ATF4 (Table S1).

We began by determining the contribution of PERK and its effectors to the transcriptional injury response. Using a stringent statistical assessment for differentially expressed genes (DEGs, false discovery rate FDR<0.05), we identified 282 transcripts modulated by injury in control wildtype mice injected with AAV2-hSyn1-Cre. Among these, 117, or 41.5%, reached this same strict threshold for differential expression in a comparison between PERK cKO retinae after injury wildtype retinae after injury (Table S1), with this number increased to 157, or 55.7%, using a moderately relaxed threshold of p<0.01 and FDR<0.2 (Figure 2A). All but three of these change in the opposite direction of their modulation by injury, suggesting a central role for the ISR in mediating these injury responses. Among the 154 transcripts regulated by injury in a PERK-dependent manner, 56 (36.4%) are also suppressed by ATF4 knockout, with a particularly pronounced contribution among those transcripts that are upregulated after injury (37 out of 73, or 50.7%), including seven of the nine transcripts that display CHOP-dependence at a threshold of p<0.01 and FDR<0.2 (Figure 2A). Notably, among ATF4-regulated transcripts, 86.2% also exhibit PERK-dependence at this threshold, consistent with ATF4 activation after optic nerve crush being mediated predominantly by PERK and not subject to compensation by other eIF2α kinases (Larhammar et al., 2017). This analysis, though limited by its arbitrary thresholds for defining factor-dependence, implies an essential role of the PERK-activated ISR in the transcriptional injury response, with CHOP serving as a relatively minor contributor within a program primarily mediated by ATF4.

The PERK-activated ISR is a prominent contributor to the transcriptional response to optic nerve injury, primarily acting through ATF4.

(A) Venn diagrams of 282 transcripts exhibiting significant differences by RNA-seq between uninjured retinae and 3 days after optic nerve crush (FDR<0.05). Amongst those, transcripts that exhibit differences between wildtype and cKO retinae after injury (p<0.01 and FDR<0.2) were classified as PERK-, ATF4-, or CHOP-regulated. (B-D) Linear regression analyses comparing the impact of neuronal PERK deletion (Eif2ak3 cKO), ATF4 deletion (Atf4 cKO), and CHOP deletion (Ddit3 cKO) on the same 282 injury-responsive transcripts. Names of representative ATF4-dependent transcripts are burnt orange text and representative CHOP-dependent transcripts are gold text. Significant changes are indicated by colored dots (FDR<0.05).

To evaluate these relationships in a threshold-independent manner, we performed linear regression of the 282 injury-regulated genes. We first compared the impact of PERK cKO on the regulation of these transcripts by injury with that of ATF4 cKO. This analysis reveals a strong correlation (R=0.87) that, along with a best-fit slope of 0.75, suggests that the role of ATF4 in mediating the effects of PERK is more comprehensive than suggested by threshold-dependent analyses alone (Figure 2B). A similar assessment uncovers a milder (R=0.59), but still highly significant, correlation between PERK-dependent and CHOP-dependent injury-responsive transcripts (Figure 2C). However, the shallow slope of 0.23 for the best-fit line reveals that, though CHOP influences similar genes as PERK, it does so relatively mildly. Consistent with these findings, we observe a significant correlation between ATF4-dependent and CHOP-dependent transcripts, again with a shallow slope that indicates a more potent influence of ATF4 knockout (Figure 2D). These results imply that CHOP serves as a potentiator of the ISR after optic nerve crush rather than a primary mediator. Together, these findings are consistent with ATF4 exerting the dominant role in the transcriptional response and promotion of RGC apoptosis.

PERK signaling regulates canonical ATF4 and c-Jun transcriptional programs

To explore the transcriptional programs downstream of PERK, we next applied the Upstream Regulator tool of Ingenuity Pathway Analysis (IPA), which includes determination of potential transcription factors likely to be mediators of the observed patterns of transcriptional changes and known target genes derived from extensive curation of the literature. Consistent with previous studies (Syc-Mazurek et al., 2022; Yasuda et al., 2016), we find that ATF4, along with c-Jun, emerges among the top five candidates for a positive transcriptional regulator (p<10-13 and Activation Z-score>2.5), based on the upregulation of many genes previously demonstrated to be direct targets of ATF4, including Atf3, Mthfd2, Stc2, and Phgdh (Ben-Sahra et al., 2016; Han et al., 2013; Pan et al., 2007; Zhao et al., 2016) (Figure 3A). IPA also suggested PGC1α (gene name: Ppargc1a), which can be negatively regulated by ATF4 activation (Montori-Grau et al., 2022; Wang et al., 2013), as the most influential negative transcriptional regulator (Table S2). CHOP (gene name: Ddit3) was implicated among a cluster of dozens of less influential positive regulators (p<10-5 and Activation Z-score>1.25) (Figure 3A; Table S2). Comparisons between wildtype and PERK (Eif2ak3) cKO retinae 3 days after injury suggest that disruption of PERK abrogates ATF4 activation most strongly, with c-Jun and PGC1α also affected (Figure 3B; Table S3). This analysis suggests the activation of a canonical PERK/ATF4-mediated transcriptional response (Figure 3C) after RGC axon injury and raised the unexpected possibility that PERK signaling might also influence c-Jun-regulated target genes.

PERK regulates transcriptional changes through canonical ATF4 target genes and influence on c-Jun-regulated programs.

(A-B) Volcano plots of Ingenuity Pathway Analysis (IPA) implicate ATF4 and c-Jun as potential Upstream Transcriptional Regulators of injury-induced expression changes in (A) wildtype retina, and (B) those whose activity after injury is reduced by neuronal PERK deletion. Red, orange, blue, and light blue dots indicate Z>2.5, p<10-13; Z>1.5, p<10-10; Z<-2.5, p<10-13; and Z<-1.5, p<10-10, respectively. (C) Heat map of RNA-seq data, showing known and putative ATF4 target genes after injury in wildtype or cKO retinae. (D-F) Linear regression analyses comparing current whole retina RNA-seq cKO data to published data from a similarly designed experiment (Syc-Mazurek et al, 2022). Data from that independent study is indicated on the y-axis by &. (D) Strong correlation among 282 injury-responsive transcripts in the respective wildtype injury conditions between these two independent studies. Names of representative ATF4-dependent transcripts are burnt orange text and representative CHOP-dependent transcripts are gold text. (E) Significant correlations between the impacts of neuronal PERK conditional knockout (this study) and deletion of c-Jun from the majority of neural retina& (p<0.001). (F) Cross-study comparison of CHOP (Ddit3) cKO (this study) and CHOP (Ddit3) KO& shows strong correlation between the few CHOP-dependent injury-responsive transcripts (red dots).

To further investigate a potential interaction between c-Jun- and PERK-regulated transcripts, we leveraged RNA-seq datasets from a separate, similarly designed, study that included mice lacking c-Jun in the majority of neural retina and CHOP-null mice (Syc-Mazurek et al., 2022). First, we evaluated the expression profiles of the wildtype injury conditions between these independent studies, finding an exceptional correlation (R=0.95, Slope = 1.07) among 282 injury-regulated genes (Figure 3D). That robust relationship between the control conditions provided a basis for further comparisons. We therefore proceeded with a cross-study analysis comparing the impact of PERK cKO with that of c-Jun cKO. That analysis revealed a significant correlation (R=0.49; Slope = 0.44) among c-Jun- and PERK-modulated transcripts (Figure 3E). Nevertheless, some injury-modulated transcripts are regulated exclusively by c-Jun (e.g., Sox11 and Arid5a) or PERK (e.g., Chac1 and Phgdh). Together, these findings suggest that, despite a lack of influence of c-Jun knockout on PERK (Eif2ak3), ATF4 (Atf4), or CHOP (Ddit3) transcripts and vice versa, these parallel pathways participate in substantially overlapping transcriptional programs.

We next evaluated the similarity between germline and neuronal knockout of CHOP (Ddit3) between these two studies. Though each uncovered only a small number of CHOP-dependent, injury-responsive transcripts, those few exhibit considerable overlap. Just four transcripts passed a stringent test for CHOP-dependence in retinae of CHOP/Ddit3-/- mice (Syc-Mazurek et al., 2022), three of which – Stbd1, Gm13889, and Avil (also known as Doc6 for “dependent on CHOP-6”) – have been demonstrated to be upregulated neuron-autonomously by injury within RGCs by scRNA-seq (Tran et al., 2019). Using similar stringency (FDR<0.05), these three were among only five genes that we independently detected as injury-upregulated in a neuronally CHOP-dependent manner using CHOP cKO mice (Figure 3F). Additional injury-induced transcripts suppressed by greater than 50% in both CHOP cKO and CHOP KO (though without surpassing the FDR<0.05 threshold) include Fibin (p<0.0005 in both studies) and Cdsn (p<0.001 in both studies), genes for which ChIP-seq has previously identified CHOP binding near their transcriptional start sites under stress conditions (Han et al. 2013). These results show that, despite differences in neuroprotection between CHOP-null mice (Hu et al., 2012; Syc-Mazurek et al., 2017b) and neuronal targeting of CHOP in cKO retinae, their transcriptional consequences within the retina closely resemble one another.

Together, these expression profiles suggest that the PERK-mediated ISR is an integral contributor to the retinal transcriptional response after optic nerve injury, acting primarily through canonical ATF4 targets and interacting with c-Jun-mediated transcriptional programs.

RGC-autonomous ATF4- and CHOP-dependent transcriptional changes are prominently represented by whole retina transcriptomics

Our targeted expression of Cre recombinase in cKO mice argues for a central role for the neuronal ISR in the response to optic axon injury, with whole retina transcriptomics likely reporting both these neuron-autonomous effects and secondary consequences to other retinal cells. To determine the extent to which ISR-dependent transcription uncovered by retinal RNA-seq represents RGC-autonomous expression changes, we next leveraged an independent single-cell RNA-seq (scRNA-seq) data set that details the injury-induced expression changes within subtypes of RGCs at 2 and 4 days after optic nerve crush (Tran et al., 2019).

We began by assessing the whole retina expression of transcripts that have been demonstrated to be differentially expressed by RGCs at 2 or 4 days after optic nerve crush (pseudo-bulk scRNA-seq |Diff>0.3| and FDR<10-20). Of 615 such transcripts, 597 are represented in wildtype whole retina 3 days after injury. We find that 290 of these transcripts exhibit differential expression (|log2FC|>0.25) in retina, with all but three modulated by injury in the same direction as reported by scRNA-seq (Figure 4A). Moreover, the majority of the 282 injury-responsive transcripts we have identified in whole retina exhibit concordance with changes previously found by scRNA-seq of RGCs (Supplemental Figure S2A-C). Importantly, many transcripts identified as PERK-, ATF4-, and/or CHOP-dependent in this and other studies are among those that are upregulated neuron-autonomously within multiple subtypes of RGCs following optic nerve crush (Figure 4B). Accordingly, RNAScope confirms that the ATF4-dependent transcripts Chac1, Ecel1, and Atf3 are elevated in the GCL in conjunction with downregulation of the RGC marker Rbpms, with these changes suppressed in ATF4 cKO retinae (Figure 4C-D). Furthermore, RNAScope corroborates that the Avil and Ecel1 transcripts are upregulated in subsets of RGCs in an ATF4- and CHOP-dependent manner (Figure 4D-E and Supplemental Figure S2D). Consistent with this, the relationships we detected between PERK-dependent transcripts and ATF4- or c-Jun-dependent transcripts in whole retina are maintained when re-assessed using DEGs confirmed to be injury-regulated within RGCs by scRNA-seq (Supplemental Figure S2E-F). These results imply that the injury-induced ATF4 transcriptional program that we detect by whole retina transcriptomics of cKO mice primarily reflects RGC-autonomous expression changes.

Cell-autonomous expression of ATF4- and CHOP-dependent transcripts by RGCs.

(A) Volcano plot of retinal expression data three days after optic nerve crush (this study) for 597 transcripts that are significantly altered autonomously in RGCs at day 2 or day 4 after crush, as determined by scRNA-seq (Tran et al., 2019). 287 of these transcripts are detected at FDR<0.05 (red=upregulated, or blue=downregulated) or |log2FC|>0.25 (light red=upregulated, or light blue=downregulated) in whole retina. Only three transcripts (yellow dots) were significantly regulated in whole retina in the discordant direction from scRNA-seq findings. (H) scRNA-seq data set (Tran et al., 2019) reveals RGC-autonomous activation of canonical ISR transcripts. Dot plot at 2 and 4 days after injury, showing both pan-RGC and type-specific upregulation of transcripts demonstrated by whole retina RNA-seq to be ATF4- and/or CHOP-dependent. Some transcripts exhibiting low, non-uniform, injury induction in RGCs – Avil, Gm29374, Cdsn, and Fibin – were amongst those that appear to exhibit potential CHOP-dependence but failed to reach threshold for inclusion in either retinal injury-dependence, CHOP-dependence, or scRNA-seq pseudo-bulk analyses. (C-E) Fluorescent in situ hybridization (RNAScope) of ATF4- and CHOP-dependent transcripts. (C) Box-and-whisker plots of whole retina transcriptomics for selected ATF4- and CHOP-dependent transcripts and the RGC marker gene Rbpms that were probed by RNAScope of fresh-frozen retinal cyrosections. (D-E) Multiplex RNAScope across wildtype (WT) uninjured (“uninj”) and 3 days post-crush (“3d”) for three genotypes (WT, ATF4 cKO, and CHOP cKO) demonstrating prominent expression by Rbpms+ RGCs of injury-induced transcripts that are reduced by knockout of ATF4 (Atf3, Chac1) or by knockout of either ATF4 or CHOP (Ecel1, Avil), concordant with RNA-seq findings. Nuclei are labeled with DAPI.

This canonical ATF4 program, inclusive of the contribution of CHOP, contrasts with a report of two non-canonical, largely non-overlapping, programs controlled by these two ISR transcription factors (Tian et al., 2022). We next investigated potential sources of this discordance. The proposed parallel neurodegenerative programs were deduced in part by RNA-seq of FACS-enriched injured RGCs expressing gRNAs targeting ATF3 or Ddit3/CHOP. We therefore began with an assessment of how the injury responses detected in whole retina in the current study compare with those reported for RGCs collected by FACS. Using only transcripts independently identified as RGC-autonomous expression changes by scRNA-seq (Tran et al. 2019) for this cross-study comparison, we find a highly significant correlation (R=0.77) between injury-regulated transcripts under control conditions (i.e., without gene targeting) between these independent studies (Supplemental Figure S3A). This suggests that these distinct approaches similarly report prominent RGC-autonomous expression changes, inclusive of numerous known direct targets of ATF4, with a steep linear regression (slope = 2.0) that is consistent with greater sensitivity for these changes by enrichment for RGCs by FACS. Despite that similarity, linear regression reveals little correlation between the impact of cKO- and gRNA-mediated knockout on 597 RGC-autonomous, injured-regulated transcripts. We find only very weak, shallow correlation between the effects of ATF4 gRNA and ATF4 cKO (R=0.25, slope = 0.52). cKO-sensitive signature ATF4 target genes Chac1, Phgdh, Sesn2, Slc7a5, and Asns, amongst others, are not among the 300 of these 598 (50.2%) transcripts reported to be significantly affected by ATF4 gRNA (Supplemental Figure S3B) (Crawford et al., 2015; Garaeva et al., 2016; Han et al., 2013; Liu et al., 2018; Oh-Hashi et al., 2013; Park et al., 2017). We also find no correlation between the influence of CHOP/Ddit3 cKO and gRNA targeting CHOP, with no reported inhibition by this gRNA pool of the CHOP KO- and cKO-sensitive transcripts Gm13889 and Stbd1 among 333 (55.7%) significantly affected transcripts (FDR<0.05) (Supplemental Figure S3C). Notably, the reported effects of gRNA targeting CHOP, like ATF3 gRNA, include minimal reduction of Ddit3/CHOP transcript but greater than 70% suppression of the pro-apoptotic and pro-regenerative Atf3 transcript (Tian et al., 2022), an effect not observed in CHOP/Ddit3 germline (Syc-Mazurek et al., 2022) or conditional knockout mice (present study), despite ample sensitivity for Atf3 mRNA modulation using whole retina transcriptomics. These analyses argue that more specific and effective disruption of target genes by cKO may contribute to the striking discordance between the transcriptional programs of ATF4 and CHOP suggested by these distinct approaches.

Neuronal ATF4 knockout limits axon regeneration by PTEN-deficient RGCs

The commonalities that we uncovered between c-Jun- and PERK-regulated transcription suggested the hypothesis that, like c-Jun, PERK/ATF4 could be involved not only in promoting apoptosis but also in promoting regeneration. To investigate this possibility, we first examined the PERK- and ATF4-dependence of regeneration-associated genes (RAGs) and the injury-induced downregulation of mature and subtype-specific RGC markers, finding that many exhibit at least partial PERK- and ATF4-dependence (Figure 5A-B). We therefore crossed mice harboring the floxxed PTEN alleles and those harboring floxxed ATF4 alleles or floxxed PERK alleles to generate homozygous dcKO mice. Following intravitreal injection of AAV2-hSyn1-Cre, we performed optic nerve crush, comparing axon regeneration in these dcKO mice to that of PTEN cKO mice (Figure 5C-D). We find that neuronal PERK deficiency or ATF4 deficiency limits the efficacy of this regenerative intervention. These data support an integral contribution of the ISR to the axon regenerative program in these CNS neurons, and indicate that, as with DLK and c-Jun, ATF4 is involved in both pro-regenerative and pro-apoptotic transcriptional changes.

PERK-ATF4 contributes to RGC axon regenerative potential after optic nerve crush.

(A-B) Heat maps of select injury-responsive transcripts revealed by retinal RNA-seq three days after optic nerve crush in wildtype and cKO mice, focusing on (A) genes implicated in axon regeneration, and (B) the maintenance of mature and subtype-specific RGC phenotypes. (C-D) Double conditional knockout (dcKO) of neuronal PERK (Eif2ak3; n=4) or ATF4 (Atf4; n=7) reduces regeneration enabled compared to conditional knockout (cKO) of the tumor suppressor PTEN alone (n=8). Error bars are SEM. Statistical analysis: two-way ANOVA with Sidak’s multiple comparisons test (**p<0.01).

Discussion

Cellular stress signaling pathways typically couple the promotion of growth and repair with priming for apoptosis, a feature of stress responses that may aid in the elimination of cells that are irretrievably damaged (Hotamisligil and Davis, 2016). As prominent effectors of stress signaling, transcription factors of the bZIP family, including c-Jun and ATF3, therefore can contribute either to apoptosis or to recovery, depending on context, complicating efforts to harness transcriptional programs for axon regeneration independently of neurodegenerative programs (Jacobi et al., 2022; Kole et al., 2020; Raivich et al., 2004; Simon and Watkins, 2018; Syc-Mazurek et al., 2017b; Tian et al., 2022; Watkins et al., 2013). Multiple lines of evidence indicating that the PERK-activated ISR contributes to RGC apoptosis after optic nerve crush raised the appealing possibility that this pathway may represent a distinct branch of the injury response that might be targeted to selectively reduce neurodegeneration without suppressing repair programs (Hu et al., 2012; Larhammar et al., 2017; Tian et al., 2022; Wang et al., 2020). The cKO studies reported here, however, reveal that the PERK-activated ISR is more integral to the entire RGC injury response than previously appreciated, contributing substantially, in conjunction with c-Jun, to both apoptotic and regenerative programs. Though MAP kinase stress signaling and the ISR are known to exhibit some degree of crosstalk in other settings (Brown et al., 2016; Danzi et al., 2018; Pakos-Zebrucka et al., 2016), the exceptionally close coupling of these largely independent pathways by DLK signaling seems to be a critical feature of the response of RGCs to axon injury, with disruption of ATF4, like that of c-Jun, limiting both neurodegeneration and axon regenerative potential.

By examining in parallel the neuronal knockout of PERK and its effectors ATF4 and CHOP, along with comparisons to related studies, the current work has provided insights into the mechanisms by which the ISR influences the fates of injured RGCs. These conditional knockout mouse experiments reveal that ATF4 is the principal effector of the ISR transcriptional response, with CHOP playing a modest role within a canonical ATF4-mediated program. Though RGC neuroprotection in germline CHOP knockout mice has long been interpreted as evidence of its central role in the neuronal ISR, our findings suggest that further investigations of CHOP may instead uncover its non-autonomous mechanisms regulating the survival of injured RGCs. The unexpectedly subtle role for neuronal CHOP aligns well with expression profiling data from CHOP-null mice (Syc-Mazurek et al., 2022) and its modest role in this study in determining RGC survival, but it appears to contrast with the more significant role for CHOP, additive with ATF4, suggested by other studies (Fang et al., 2023; Tian et al., 2022). One factor that may influence the results of CHOP cKO experiments is the ∼2-fold elevated expression of a collection of eight genes (B4galnt1, Slc26a10, Arhgef25, Dtx3, F420014N23Rik, Pip4k2c, Kif5a, and Dctn2) near the Ddit3/CHOP locus on chromosome 10. One of those genes, Kif5a, is injury-responsive and alterations to its expression can influence RGC survival (Shah et al., 2022). This confound may be expected to have also been present in an independent study using the same mouse model and is therefore not likely to account for the reported differences in RGC neuroprotection (Fang et al., 2023). Though the apparent reduction of ATF3 reported for gRNA targeting Ddit3/CHOP, not seen in CHOP KO or cKO models, may suggest an explanation for its strikingly similar reported effects of ATF3 gRNA on the transcriptome and RGC survival, those results seem incongruent with the differential reported effects of ATF3 and CHOP gRNA on RGC axon regeneration (Jacobi et al., 2022; Tian et al., 2022). With cKO experiments revealing instead that ATF3 upregulation is dependent on ATF4 and c-Jun, it will be of interest to elucidate how much of the commonality among the transcriptional programs of these bZIP factors is attributable to interdependence of their expression and how much to their combinatorial heterodimerization. Among other non-exclusive possibilities, the commonalities between ATF4- and c-Jun-dependent transcription could reflect: (1) dependence of ATF3 elevation on c-Jun-ATF4 heterodimers; (2) dependence on ATF4 for induction of ATF3, which then heterodimerizes with c-Jun; or (3) regulation of common target genes through distinct DNA binding sites. The current study provides a framework for deciphering these networks in part by leveraging cross-study comparisons of retina, RGC, and single-cell RNA-seq data to uncover robust RGC-autonomous expression signatures of Atf4 and other transcription factors.

Given the potential for the ISR to influence neuronal fate through multiple translational and transcriptional mechanisms, it is perhaps surprising that its primary contributions to the fates of injured RGCs are mediated by a single transcription factor, ATF4. Nevertheless, at least four lines of evidence contend that the cKO experiments reported here provide a reliable picture of its contribution. First, the extensive commonality in injury-regulated transcripts across studies, whether utilizing retinal expression data or filtering for RGC-autonomous expression changes, argues against AAV2-hSyn1-Cre or other technical aspects of our approach altering the injury response. Secondly, support for the modest impact of CHOP disruption on RGC injury-regulated genes is provided by the highly similar findings in the independently generated and validated CHOP-null and CHOP cKO mouse lines. Third, we find remarkable overlap between expression changes and phenotypes mediated by PERK and ATF4, which we previously demonstrated to be in the same pathway after optic nerve crush and other DLK-activating insults (Larhammar et al., 2017). Finally, the interaction between PERK/ATF4 and c-Jun-regulated transcription uncovered by our cross-study analysis predicts an impact of PERK or ATF4 knockout on RGC axon regenerative potential that is validated by dcKO with PTEN. This result is consistent with recent findings that early injury responses, including regenerative and degenerative responses that we have found in this study to be driven by PERK/ATF4, are similar in non-regenerating wildtype and regenerating PTEN-deficient RGCs (Jacobi et al, 2022). Together, these conditional knockout studies highlight the critical roles for ISR-activated transcriptional programs in determining the fates of distressed neurons and the intimate link between injury-activated neurodegenerative and axon regenerative programs.

Supporting information